Stack Safety is a Security Property

### 2021-04-16 tex/main.tex
Polishing

Colored boxes indicate an assignment of the address to a
{\em domain} -- a label reflecting a particular set of access rights. Domains
are \(\unsealed\)\bcp{``unsealed'' would read better} (in the stack and accessible to writes), \(\sealed{\depth}\)
(in the stack and reserved for the caller at depth \(\depth\)), and \(\outside\)
(outside the stack entirely.) For our purposes, \(\outside\) will apply to global
variables. Domains are part of the {\em context} of a state.
We will introduce the remaining context as needed.

@@

Each system component is assigned to a {\em domain}, which may be {\em Outside}
of the stack entirely, {\em Unsealed} and therefore accessible, or {\em Sealed}
at a particular depth. Each caller registers a {\em target}
condition -- a predicate on machine states that must hold for it to consider itself active.
A context is a pair of a ``domain map'' from components to domains, and a
``return target list,'' which is a stack of targets, one for each caller awaiting
return.  \ifaftersubmission\bcp{spacing is funny}\fi
%
\[\stackDom \in \STKDOMS ::= \outside + \unsealed + \sealed{\depth} \hspace*{0.5in} \target \subseteq \MACHS\]
\[(\domMap,\rts) = \context \in \CONTEXTS ::= (\COMPONENTS \rightarrow \STKDOMS)
  \times (\listT (\target)) \]
%
The initial context \(\context_0 = \domMap_0, []\), where \(\domMap_0\) maps
the stack to \(\unsealed\) and all other addresses and all registers to \(\outside\).

@@

\definition {\bf Stepwise Stack Integrity}

A system enjoys {\em stepwise stack integrity} if, for any reachable state
\(\mpcstatename\) such that \(\mpcstatename \stepstocon \mpcstatename'\),
and any component \(\component\), if \(\pi_\context(\mpcstatename) = \domMap,\_\) and
\(\domMap ~ \component = \sealed{\_}\), then \(\pi_\mach(\mpcstatename') ~ \component =
\pi_\mach(\mpcstatename) ~ \component\).

@@

\definition {\bf Stepwise Stack Confidentiality}

For any call trace \(\MPCT \in \MPCT^*\) with \(\head(\MPCT) = \mach,\pol,(\domMap,\rts)\),
let \(\components\) be the set of components \(\component\) such that
\(\domMap ~ \component = \outside\). That is, at the start of some call, we take every
component in the stack.
A system enjoys \emph{stepwise stack confidentiality} if,
for all \(\nach\) such that \(\mach \approx_\components \nach\) and
\(\nach,\pol,\domMap \hookrightarrow \NPCT | \neg f_{|\rts|}\),
\(\MPCT \doteq \NPCT\) -- if every variant (with the same policy state)
steps in lockstep until both traces have reached a point whose depth is
below the starting one. This is a return.

\vspace{\belowdisplayskip}

@@

\definition {\bf Return-To-Caller}

A system enjoys {\em return-to-caller} when, for every pair of reachable states
\(\mpcstatename = (\_,\_,(\_,\rts))\) and
\(\mpcstatename' = (\_,\_,(\_,\rts'))\) such that
\(\mpcstatename \stepsto \mpcstatename'\), if \(|\rts'| < |\rts|\), then
\(|\rts'| = |\rts|\)

@@

Next we introduce variable passing. We will need to add extra information to our call
annotations to reflect that they may now pass variables on the stack, and that the locations
of those variables may be dynamic (as in the case of pass-by-reference).
A call annotation now includes a predicate relating machine states to the components
that should be passed -- when passing by value, this will typically be freshly allocated
cells at the top of the frame, while passing by reference just gives the
addresses of the appropriate variables.
\[\codemap \in \CODEMAPS ::= \addr \in \WORDS \rightharpoonup \callmap(P \subseteq \MACHS \times \COMPONENTS)\]
And we extend the set of domains to include passed components,
which are labeled with the range of depths of the caller that has passed them.
\[\stackDom \in \STKDOMS ::= \outside + \unsealed + \sealed{\depth} + \passed{\depth,\depth'}\]
\[\context \in \CONTEXTS ::= (\domMap \in \COMPONENTS \rightarrow \STKDOMS)
\times (\rts \in \listT (\target \subseteq \MACHS)) \]
We straightforwardly describe how the sealing convention and passing predicate combine to
update a domain map when pushing it, where \(\depth\) is an integer depth.
\[(\mathit{push} ~ \domMap ~ \sealcon ~ P ~ \mach ~ \depth) ~ \component =
\begin{cases}
  \sealed{\depth}              & \text{if } \domMap ~ \component = \unsealed \text{ and }
                                 \sealcon ~ \mach ~ \component \text{ and } \neg P ~ \mach ~ \component \\
  \passed{\depth,\depth+1}     & \text{if } \domMap ~ \component = \unsealed \text{ and }
                                 \sealcon ~ \mach ~ \component \text{ and } P ~ \mach ~ \component \\
  \passed{\depth',\depth+1}    & \text{if } \domMap ~ \component = \passed{\depth',\depth} \text{ and }
                                 \sealcon ~ \mach ~ \component \text{ and } P ~ \mach ~ \component,
                                 \text{ for any } \depth' \\
  \domMap~\component & \text{else} \\
\end{cases}\]

@@

\definition {\bf Stepwise Stack Integrity (Passing)}

Let \(\mpcstatename\) and \(\mpcstatename'\) be reachable states such that
\(\mpcstatename \stepstocon \mpcstatename'\), and \(\pi_\context(\mpcstatename) = \domMap,\rts\).
A system enjoys {\em stepwise stack integrity with passing} if for all
such states, \(\pi_\mach(\mpcstatename) ~ \component = \pi_\mach(\mpcstatename') ~ \component\)
for each component in \(\mathit{protected}(\pi_\context(\mpcstatename))\)

\definition {\bf Stepwise Stack Confidentiality (Passing)}

For any call trace \(\MPCT \in \MPCT^*\) with \(\head(\MPCT) = \mach,\pol,(\domMap,\rts)\),
let \(\components\) be the set of components \(\component\) such that
\(\domMap ~ \component = \outside\) or \(\domMap ~ \component = \passed{\depth,\depth'}\)
and \(\depth \leq |\rts| \leq \depth'\). A system enjoys
\emph{stepwise stack confidentiality} if,
for all \(\nach\) such that \(\mach \approx_\components \nach\) and
\(\nach \hookrightarrow \NPCT | \neg f_{|\rts|}\),
\(\MPCT \doteq \NPCT\).

@@

This discipline suffices to enforce local stack encapsulation.
To implement it,  we define ``blessed'' instruction sequences
intended to appear at the entry and exit of each function,
which manipulate tags as just described in addition to performing the
usual calling convention tasks of saving/restoring the return address to/from
the stack and adjusting the stack pointer. With the aid of an additional tag on
the $\rsp$ register, these sequences also serve to enforce
well-bracketed control flow.
The micro-policy guarantees atomic execution of these sequences
using a combination of tags on the instructions
and an additional tag on the $\PCname$; we omit the details here.

@@

\definition {\bf Deferred Stack Integrity}

For any \(\MPCT \in \MPCT^*\),
let \(\head(\MPCT) = \mach,\pol,(\domMap,\rts)\) and
\(\components\) be the set of components \(\component\) such that
\(\domMap ~ \component = \sealed{\_}\).
Then if \(\last(\MPCT) = \mach',\pol',\context'\), let \(\components' = \Delta(\mach,\mach')\).
 A system enjoys {\em deferred stepwise integrity} if for any \(\nach\) such that
\(\mach' \approx_{\components'} \nach\), \(\mach \hookrightarrow \MPCT\), and
\(\nach \hookrightarrow \NPCT\), \(\MPCT \doteq \NPCT\).

\vspace{\belowdisplayskip}

@@

So we extend our model with a notion of observations. We abstract over an
observation type \(\obs \in \OBSS\), containing at minimum the silent observation
\(\tau\). Our machine step function now takes a state and returns a state and an observation,
which is carried over into the policy and context step functions.
%
\[\mach \stepsto[\obs] \mach' \in \MACHS \rightarrow \MACHS \times \OBSS \]
%

@@

We now can sketch observational versions of both integrity and confidentiality.
Integrity is already straightforward, simply weakening the condition that must hold
after the return -- instead of preventing dangerous writes from being read, we
prevent them from being leaked.

\definition {\bf Observational Stack Integrity}

For any call trace \(\MPCT \in \MPCT^*\),
let \(\head(\MPCT) = \mach,\pol,(\domMap,\rts)\) and
\(\components\) be the set of components \(\component\)
such that \(\domMap ~ \component = \sealed{\_}\).

Then, supposing that \(\last(\MPCT) = \mach',\pol',\context'\),
let \(\components'\) be the set of components that differ
between \(\mach\) and \(\mach'\). A system enjoys
{\em observational integrity} if for any \(\nach\) such that
\(\mach' \approx_{\components'} \nach\), \((\mach,\pol',\context') \hookrightarrow \MPCT\), and
\((\nach,\pol',\context') \hookrightarrow \NPCT\), \(\obsof(\MPCT) \simeq \obsof(\NPCT)\).

\vspace{\belowdisplayskip}

@@

\definition {\bf Observational Stack Confidentiality}

For any call trace \(\MPCT \in \MPCT^*\),
let \(\head(\MPCT) = \mach,\pol,(\domMap,\rts)\) and
\(\components\) be the set of components \(\component\)
such that \(\domMap ~ \component \not = \outside\).

A system enjoys {\em observational confidentiality} if for all
\(\nach\) such that \(\mach \approx_\components \nach\) and
\(\nach \hookrightarrow \NPCT | \neg f_\depth \),
\(\MPCT\) and \(\NPCT\), the following conditions all hold:

\begin{itemize}
\item They are observationally equivalent, \(\obsof(\MPCT) = \obsof(\NPCT)\),
  meaning that no secrets leak during the call
\item \(f_\depth ~ \last(\MPCT) \leftrightarrow f_\depth ~ \last(\NPCT)\) --
  if one call returns, so does the other
\item Assuming both calls return, let \(\last(\MPCT) = \mach',\pol',\context'\)
  and \(\pi_\mach(\last(\NPCT)) = \nach'\).
  Let \(\components' = \bar{\Diamond}(\components,\mach,\mach',\nach,\nach'\).
  And let \(\nach''\) be a state such that \(\mach' \approx_\components \nach''\).
  Then if \(\mach',\pol',\context' \hookrightarrow \MPCT'\) and
  \(\nach',\pol',\context' \hookrightarrow \NPCT'\), it must be
  that \(\obsof(\MPCT') \simeq \obsof(\NPCT')\)
\end{itemize}

@@

Lazy Per-activation Tagging and Clearing ends up enforcing the combination
of deferred stepwise stack integrity and stepwise confidentiality -- we
don't actually need a deferred confidentiality property here. Therefore,
it also enforces observational stack safety, so we declare it stack safe.
Our properties have allowed us to reject an enforcement mechanism as
not really stack safe, and describe a fix.

@@

There are several proposals around the use of Cheri capabilities to enforce stack safety,
ranging from mechanisms that use the standard Cheri hardware \citep{Skorstengaard+19b},
to those that propose entirely new types of capabilities, such as linear
(\citep{Skorstengaard+19}), uninitialized (\citep{Georges+21}), and monotonic (\sna{cite}).
Of these, uninitialized capabilities seem particularly promising -- they naturally enforce
write-before-read behavior, and therefore will enable confidentiality without the need
to clear large swaths of the stack. We propose to implement a model of the stack safety
system described in (\citep{Georges+21}) and test it against our properties using
QuickChick.

>>>>

Colored boxes indicate an assignment of the address to a
{\em domain}---a label reflecting a particular set of access rights. Domains
are \(\unsealed\)\bcp{``unsealed'' would read better} (in the stack and accessible to writes), \(\sealed{\depth}\)
(in the stack and reserved for the caller at depth \(\depth\)), and \(\outside\)
(outside the stack entirely.) For our purposes, \(\outside\) will apply to global
variables. Domains are part of the {\em context} of a state.
We will introduce the remaining context as needed.

@@

Each system component is assigned to a {\em domain}, which may be {\em Outside}
of the stack entirely, {\em Unsealed} and therefore accessible, or {\em Sealed}
at a particular depth. Each caller registers a {\em target}
condition---a predicate on machine states that must hold for it to consider itself active.
A context is a pair of a ``domain map'' from components to domains, and a
``return target list,'' which is a stack of targets, one for each caller awaiting
return.  \ifaftersubmission\bcp{spacing is funny}\fi
%
\[\stackDom \in \STKDOMS ::= \outside + \unsealed + \sealed{\depth} \hspace*{0.5in} \target \subseteq \MACHS\]
\[(\domMap,\rts) = \context \in \CONTEXTS ::= (\COMPONENTS \rightarrow \STKDOMS)
  \times (\listT (\target)) \]
%
The initial context \(\context_0 = \domMap_0, []\), where \(\domMap_0\) maps
the stack to \(\unsealed\) and all other addresses and all registers to \(\outside\).

@@

\definition
A system enjoys {\em stepwise stack integrity} if, for any reachable state
\(\mpcstatename\) such that \(\mpcstatename \stepstocon \mpcstatename'\),
and any component \(\component\), if \(\pi_\context(\mpcstatename) = \domMap,\_\) and
\(\domMap ~ \component = \sealed{\_}\), then \(\pi_\mach(\mpcstatename') ~ \component =
\pi_\mach(\mpcstatename) ~ \component\).

@@

\definition
For any call trace \(\MPCT \in \MPCT^*\) with \(\head(\MPCT) = \mach,\pol,(\domMap,\rts)\),
let \(\components\) be the set of components \(\component\) such that
\(\domMap ~ \component = \outside\). That is, at the start of some call, we take every
component in the stack.
A system enjoys \emph{stepwise stack confidentiality} if,
for all \(\nach\) such that \(\mach \approx_\components \nach\) and
\(\nach,\pol,\domMap \hookrightarrow \NPCT | \neg f_{|\rts|}\),
\(\MPCT \doteq \NPCT\)---if every variant (with the same policy state)
steps in lockstep until both traces have reached a point whose depth is
below the starting one. This is a return.

\medskip

@@

\definition
A system enjoys {\em return-to-caller} when, for every pair of reachable states
\(\mpcstatename = (\_,\_,(\_,\rts))\) and
\(\mpcstatename' = (\_,\_,(\_,\rts'))\) such that
\(\mpcstatename \stepsto \mpcstatename'\), if \(|\rts'| < |\rts|\), then
\(|\rts'| = |\rts|\)

@@

Next we introduce variable passing. We will need to add extra information to our call
annotations to reflect that they may now pass variables on the stack, and that the locations
of those variables may be dynamic (as in the case of pass-by-reference).
A call annotation now includes a predicate relating machine states to the components
that should be passed---when passing by value, this will typically be freshly allocated
cells at the top of the frame, while passing by reference just gives the
addresses of the appropriate variables.
\[\codemap \in \CODEMAPS ::= \addr \in \WORDS \rightharpoonup \callmap(P \subseteq \MACHS \times \COMPONENTS)\]
And we extend the set of domains to include passed components,
which are labeled with the range of depths of the caller that has passed them.
\[\stackDom \in \STKDOMS ::= \outside + \unsealed + \sealed{\depth} + \passed{\depth,\depth'}\]
\[\context \in \CONTEXTS ::= (\domMap \in \COMPONENTS \rightarrow \STKDOMS)
\times (\rts \in \listT (\target \subseteq \MACHS)) \]
We straightforwardly describe how the sealing convention and passing predicate combine to
update a domain map when pushing it, where \(\depth\) is an integer depth.
\[(\mathit{push} ~ \domMap ~ \sealcon ~ P ~ \mach ~ \depth) ~ \component =
\begin{cases}
  \sealed{\depth}              & \text{if } \domMap ~ \component = \unsealed \text{ and }
                                 \sealcon ~ \mach ~ \component \text{ and } \neg P ~ \mach ~ \component \\
  \passed{\depth,\depth+1}     & \text{if } \domMap ~ \component = \unsealed \text{ and }
                                 \sealcon ~ \mach ~ \component \text{ and } P ~ \mach ~ \component \\
  \passed{\depth',\depth+1}    & \text{if } \domMap ~ \component = \passed{\depth',\depth} \text{ and }
                                 \sealcon ~ \mach ~ \component \text{ and } P ~ \mach ~ \component,
                                 \text{ for any } \depth' \\
  \domMap~\component & \text{else} \\
\end{cases}\]

@@

\definition
Let \(\mpcstatename\) and \(\mpcstatename'\) be reachable states such that
\(\mpcstatename \stepstocon \mpcstatename'\), and \(\pi_\context(\mpcstatename) = \domMap,\rts\).
A system enjoys {\em stepwise stack integrity with passing} if for all
such states, \(\pi_\mach(\mpcstatename) ~ \component = \pi_\mach(\mpcstatename') ~ \component\)
for each component in \(\mathit{protected}(\pi_\context(\mpcstatename))\)

\definition
For any call trace \(\MPCT \in \MPCT^*\) with \(\head(\MPCT) = \mach,\pol,(\domMap,\rts)\),
let \(\components\) be the set of components \(\component\) such that
\(\domMap ~ \component = \outside\) or \(\domMap ~ \component = \passed{\depth,\depth'}\)
and \(\depth \leq |\rts| \leq \depth'\). A system enjoys
\emph{stepwise stack confidentiality} if,
for all \(\nach\) such that \(\mach \approx_\components \nach\) and
\(\nach \hookrightarrow \NPCT | \neg f_{|\rts|}\),
\(\MPCT \doteq \NPCT\).

@@

This discipline suffices to enforce local stack encapsulation.
To implement it,  we define ``blessed'' instruction sequences
intended to appear at the entry and exit of each function,
which manipulate tags as just described in addition to performing the
usual calling convention tasks of saving/restoring the return address to/from
the stack and adjusting the stack pointer. With the aid of an additional tag on
the $\rsp$ register, these sequences also serve to enforce
well-bracketed control flow.
The micro-policy guarantees atomic execution of these sequences
using a combination of tags on the instructions
and an additional tag on the $\PCname$.

@@

\definition
For any \(\MPCT \in \MPCT^*\),
let \(\head(\MPCT) = \mach,\pol,(\domMap,\rts)\) and
\(\components\) be the set of components \(\component\) such that
\(\domMap ~ \component = \sealed{\_}\).
Then if \(\last(\MPCT) = \mach',\pol',\context'\), let \(\components' = \Delta(\mach,\mach')\).
 A system enjoys {\em deferred stepwise integrity} if for any \(\nach\) such that
\(\mach' \approx_{\components'} \nach\), \(\mach \hookrightarrow \MPCT\), and
\(\nach \hookrightarrow \NPCT\), \(\MPCT \doteq \NPCT\).

% \vspace{\belowdisplayskip}

@@

So we extend our model with a notion of observations. We abstract over an
observation type \(\obs \in \OBSS\), containing at minimum the silent observation
\(\tau\). Our machine step function now takes a state and returns a state and an observation,
which is carried over into the policy and context step functions:
%
$\mach \stepsto[\obs] \mach' \in \MACHS \rightarrow \MACHS \times \OBSS $.
%

@@

We now can sketch observational versions of both integrity and confidentiality.
Integrity is already straightforward, simply weakening the condition that must hold
after the return---instead of preventing dangerous writes from being read, we
prevent them from being leaked.

\definition
For any call trace \(\MPCT \in \MPCT^*\),
let \(\head(\MPCT) = \mach,\pol,(\domMap,\rts)\) and
\(\components\) be the set of components \(\component\)
such that \(\domMap ~ \component = \sealed{\_}\).
%
Then, supposing that \(\last(\MPCT) = \mach',\pol',\context'\),
let \(\components'\) be the set of components that differ
between \(\mach\) and \(\mach'\). A system enjoys
{\em observational integrity} if for any \(\nach\) such that
\(\mach' \approx_{\components'} \nach\), \((\mach,\pol',\context') \hookrightarrow \MPCT\), and
\((\nach,\pol',\context') \hookrightarrow \NPCT\), \(\obsof(\MPCT) \simeq \obsof(\NPCT)\).

\medskip

@@

\definition
For any call trace \(\MPCT \in \MPCT^*\),
let \(\head(\MPCT) = \mach,\pol,(\domMap,\rts)\) and
\(\components\) be the set of components \(\component\)
such that \(\domMap ~ \component \not = \outside\).
%
A system enjoys {\em observational confidentiality} if for all
\(\nach\) such that \(\mach \approx_\components \nach\) and
\(\nach \hookrightarrow \NPCT | \neg f_\depth \),
\(\MPCT\) and \(\NPCT\), the following conditions all hold:

\begin{itemize}
\item They are observationally equivalent, \(\obsof(\MPCT) = \obsof(\NPCT)\),
  meaning that no secrets leak during the call
\item \(f_\depth ~ \last(\MPCT) \leftrightarrow f_\depth ~
\last(\NPCT)\)---if one call returns, so does the other
\item Assuming both calls return, let \(\last(\MPCT) = \mach',\pol',\context'\)
  and \(\pi_\mach(\last(\NPCT)) = \nach'\).
  Let \(\components' = \bar{\Diamond}(\components,\mach,\mach',\nach,\nach'\).
  And let \(\nach''\) be a state such that \(\mach' \approx_\components \nach''\).
  Then if \(\mach',\pol',\context' \hookrightarrow \MPCT'\) and
  \(\nach',\pol',\context' \hookrightarrow \NPCT'\), it must be
  that \(\obsof(\MPCT') \simeq \obsof(\NPCT')\)
\end{itemize}

@@

{\em Lazy Per-activation Tagging and Clearing} ends up enforcing the combination
of deferred stepwise stack integrity and stepwise confidentiality---we
don't actually need a deferred confidentiality property here. Therefore,
it also enforces observational stack safety, so we declare it stack safe.
Our properties have allowed us to reject an enforcement mechanism as
not really stack safe, and describe a fix.

@@

There are several proposals around the use of Cheri capabilities to enforce stack safety,
ranging from mechanisms that use the standard Cheri hardware \citep{Skorstengaard+19b},
to those that propose entirely new types of capabilities, such as linear
(\citep{Skorstengaard+19}), uninitialized (\citep{Georges+21}), and monotonic (\sna{cite}).
Of these, uninitialized capabilities seem particularly promising---they naturally enforce
write-before-read behavior, and therefore will enable confidentiality without the need
to clear large swaths of the stack. We propose to implement a model of the stack safety
system described in (\citep{Georges+21}) and test it against our properties using
QuickChick.
